Efficient pedigree recording for fast population genetics simulation

### 2017-11-19 forwards_paper.tex
finished simplify description

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}

@@

Here we describe the simplification algorithm.
The flow is close to our implementation,
but the description below uses set operations,
while in the implementation,
all ancestors are maintained as linked lists of segments
ordered by left endpoint.

\begin{taocpalg}{S}{Simplify a tree sequence}
{Simplifies the input tree sequence:
    input consists of 
    a list of sample IDs, $S$, of length $n$,
    a list of edges, $\mathcal{E_I}$,
    and a list of nodes $\mathcal{N_I}$.
    The genome length is $L$.
    The output will be stored in a list of edges $\mathcal{E_O}$ 
    and nodes $\mathcal{N}_O$.
    This also maintains the list $N$ that gives the mapping 
    from $\mathcal{N_I}$ to $\mathcal{N_O}$,
    so that if $M[p] = u$ then the $p^\text{th}$ ancestor in $N_I$
    is the same as the $u^\text{th}$ ancestor in $N_0$.
    The internal state is $\mathcal{A}$, the list of ancestors
    (which are collections of ancestral segments).
}

\algstep{S0.}{Initialize.}{
    Set $\mathcal{A}[j] \leftarrow (0, L, j)$ 
    and $\mathcal{N}_O[j] \leftarrow \mathcal{N}_I[S[j]]$ for $0 \le j < N$.
    Set $Q \leftarrow \emptyset$,
    and $M[p] \leftarrow -1$ for $0 \le p < \text{length}(N_I)$.
}

\algstep{S1.}{Input parent loop.}{Loop over parents, chronologically.
}

\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} 
    on each edge corresponding to the current parent.
}

\algstep{S3.}{Merge ancestors.}{Merge the resulting segments into a new ancestor for this parent.
}

@@

The first step is, for each edge $(\ell, r, p, c)$,
simply to replace any ancestry segment $[a, b)$ held by ancestor $c$
with $[a,b) \setdiff [\ell, r)$ 
(deleting the segment or splitting in two if necessary),
and adding the removed segment $[a,b) \cap [\ell, r)$ to the merge queue $Q$:
\begin{taocpalg}{R}{Remove ancestry}
    {Given an edge $(\ell, r, p, c)$,
    and a list $\mathcal{A}[c]$ of the $m$ ancestral segments corresponding to $c$,
    remove segments overlapping $[\ell, r)$ from $\mathcal{A}[c]$,
    and add those overlapping segments to $Q$.
    }

    \algstep{R1.}{Segment loop.}{Loop over ancestral segments in $\mathcal{A}[c]$:
    if $j=m$, terminate; else let $(a, b, u) = \mathcal{A}[c][j]$.
    }

    \algstep{R2.}{Add to merge queue.}{If $[a,b) \cap [\ell,r) \neq \emptyset$,
    append $(\max(a,\ell), \min(b,r), u)$ to $Q$.
    }

    \algstep{R3.}{Remove ancestry.}{If $[a,b) \cap [\ell,r) != \emptyset$,
    delete $\mathcal{A}[c][j]$ and insert in its place
    $(\min(a,\ell), \max(a,\ell), u)$ (if $a<\ell$) 
    and/or $(\min(b,r), \max(b,r), u)$ (if $r<b$).
    }

@@

Next we must merge these ancestry segments,
creating the new ancestor $p$,
outputting edges and a node if coalescence has occurred.
\begin{taocpalg}{M}{Merge ancestors}
    {Merge a collection of segments and create a new ancestor.
    This operates on the merge queue $Q$ above,
    which is a list of ancestral segments,
    and creates the new ancestor $\mathcal{A}[p]$.}

    \algstep{M1.}{Segment loop.}{While $\mathrm{length}(Q)>0$:}

    \algstep{M2.}{Find next segment overlap boundaries}{
    Let $\ell \leftarrow \min\{x.\ell : x \in Q\}$
    and $H \leftarrow \{x \in Q : x.\ell = \ell\}$.
    Also let $\ell_+ \leftarrow \min\{x.\ell : x \in Q \setdiff H\}$
    and $r = \min(\ell_+, \min\{x.r : x \in H\})$.
    }

    \algstep{M3.}{Copy single segment.}{If $\mathrm{length}(H) == 1$,
    set $\alpha \leftarrow (\ell, r, H[0].n)$.
    If also $p \in S$, then append $(\alpha.\ell, \alpha.r, N[p], \alpha.c)$ to $\mathcal{E}_O$
    and then set $\alpha.c \leftarrow M[p]$.
    Go to \algref{M7}.
    }

    \algstep{M4.}{Coalescence head.}{If $(h \leftarrow \mathrm{length}(H)) > 1$,
    if $M[p] == -1$ set $M[p] \leftarrow \text{length}(N_O)$
    and append $N_I[p]$ to $N_O$.
    Set $\alpha \leftarrow (\ell, r, M[p])$.
    }

    \algstep{M6.}{Record edges.}{
    For each $x \in H$, append $(\ell, r, M[p], x.n)$ to $\mathcal{E}_O$.
    }

    \algstep{M7.}{Update $Q$.}{
    Set $Q \leftarrow Q \setdiff \{x \in Q : x.r == r\}$
    and then set $x.\ell = \ell$ for all remaining $x \in Q$.
    }

    \algstep{M8.}{Add segment to ancestor.}{
    Append $\alpha$ to $\mathcal{A}[p]$ and return to \algref{M1}.
    }

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

Here we describe the simplification algorithm.
The flow is close to our implementation,
but the description below uses set operations,
while in the implementation,
all ancestors are maintained as linked lists of segments
ordered by left endpoint.
As it goes, the algorithm 
records a list $M$ that gives the mapping
from input nodes to output nodes,
so that if $M[p] = u$ then the $p^\text{th}$ node in the input
is the same as the $u^\text{th}$ node in the output.
The current state of the algorithm is recorded in $\Al$, 
a list of ancestors
each of which is a collections of ancestral segments.
An ancestral segment is a triple $(\ell, r, x)$,
indicating that the output node $x$ inherits from this ancestor
on the genomic segment $[\ell, r)$.

\begin{taocpalg}{S}{Simplify a tree sequence}
{Input consists of
    a list $S$ of sample IDs,
    a list $\Nt_I$ of nodes (ordered by birth time),
    a list $\Et_I$ of edges,
    and the genome length, $L$.
    The output is a 
    nodes $\Nt_O$,
    and list of edges $\Et_O$.
}

\algstep{S0.}{Initialize.}{
    Set $\Al[j] \leftarrow (0, L, j)$
    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
    Set the current input parent index to $p=0$
    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
}

\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
}

\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
}

\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
    (which adds to $\Nt_O$ and $\Et_O$).
}

\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
    otherwise, return to \algref{S1}.
}

@@

The algorithm steps back up through the pedigree,
since nodes are ordered by time-since-birth.
First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
so, for every edge $(\ell, r, p, c)$,
Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
with $[u,v) \setdiff [\ell, r)$
(deleting the segment or splitting in two if necessary),
and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:

\begin{taocpalg}{R}{Remove ancestry}
    {Given an edge $(\ell, r, p, c)$,
    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
    and add those overlapping segments to $Q$.
    }

    \algstep{R0.}{Initialize.}{
        Set $j \leftarrow 0$.
    }

    \algstep{R1.}{Segment loop.}{
    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
    (Here $x$ is the output node that inherits from this segment.)
    }

    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
    skip to \algref{R4}.
    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$. 
    }

    \algstep{R3.}{Remove ancestry.}{
    Delete $\Al[c][j]$ and insert in its place
    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
    }

    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
    }

@@

Next we must merge the ancestry segments in $Q$,
creating the new ancestor corresponding to input node $p$,
adding a node and edges to the output if any segments overlap,
which implies that an output coalescence has occurred.
Below we index elements of an ancestral segment $z = (\ell, r, u)$
by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
and $z.\lid = u$ for the output node ID.

\begin{taocpalg}{M}{Merge ancestors}
    {Merge the ancestral segments in $Q$,
    and create the new ancestor $\Al[p]$.
    This algorithm also appends to $\Nt_O$ and $\Et_O$,
    and needs to know about $S$, 
    to check if input nodes are samples.
    }

    \algstep{M0.}{Initialize.}{
        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
    }

    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}

    \algstep{M2.}{Find next segment overlap boundaries}{
        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
    }

    \algstep{M3.}{Copy single segment.}{
        If $\len(H) = 1$,
        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
        If also $p \in S$, then
        set $\alpha.\lid \leftarrow M[p]$ and
        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
        Go to \algref{M7}.
    }

    \algstep{M4.}{New output node.}{
        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
        Set $\alpha \leftarrow (\ell, r, M[p])$.
    }

    \algstep{M6.}{Record edges and update $Q$.}{
        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
    }

    \algstep{M7.}{Update $Q$.}{
        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
    }

    \algstep{M8.}{Add segment to ancestor.}{
        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
    }

    \algstep{M9.}{Defragment.}{
        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
        with the same $\lid$.  Terminate.
    }
